---
title: "TidyTuesday: Sister Cities"
date: "2026-05-12"
format:
  html:
    resources:
      - cities.csv
      - links.csv
filters:
  - shinylive
categories: ["data science", "rstats", "tidytuesday"]
---

## Introduction

This week's TidyTuesday explores the connections between sister cities across the globe. Unlike previous posts, this one is going to use shiny to create an interactive app.

In particular, I'm interested in identifying chains of sister cities that connect two countries.

### Setup

```{r setup, message=FALSE}
library(tidyverse)
library(shiny)
```

### Loading the Data

```{r load-data, message=FALSE}
# tuesdata <- tidytuesdayR::tt_load('2026-05-12')
# cities   <- tuesdata$cities |>
#   write_csv("cities.csv")
# links    <- tuesdata$links |>
#   write_csv("links.csv")

cities <- read_csv("cities.csv")
links  <- read_csv("links.csv")

```

### Chain finding algorithm 

To find the connection between two cities, we can treat the sister city relationships as an undirected graph. The following function uses the `igraph` package to find the shortest path between any two city codes.

```{r find-chain}
library(igraph)

links_graph <- graph_from_data_frame(links, directed = FALSE)

find_city_chain <- function(city_a, city_b, links_graph) {
  # Create an undirected graph from the links
  # The source and target are interchangeable
  
  # Check if cities exist in the network
  if (!(city_a %in% V(links_graph)$name) || !(city_b %in% V(links_graph)$name)) {
    return(NULL)
  }
  
  # Find the shortest path
  path_res <- shortest_paths(links_graph, from = city_a, to = city_b)
  path <- path_res$vpath[[1]]
  
  if (length(path) == 0) {
    return(NULL)
  }
  
  # Return the sequence of city codes
  return(names(path))
}
```

### Interactive Explorer

```{shinylive-r}
#| standalone: true
#| viewerHeight: 700

library(shiny)
library(leaflet)
library(igraph)
library(readr)
library(dplyr)

# Load data - in Shinylive, files must be in the same directory or listed in resources
cities <- read_csv("cities.csv")
links <- read_csv("links.csv")

# Create undirected graph for pathfinding
links_graph <- graph_from_data_frame(links, directed = FALSE)

# Prepare city choices for the dropdowns
city_choices <- cities$id
names(city_choices) <- paste0(cities$name, " (", cities$countrycd, ")")

ui <- fluidPage(
  theme = bslib::bs_theme(bootswatch = "flatly"),
  titlePanel("Sister City Chain Finder"),
  sidebarLayout(
    sidebarPanel(
      p("Find the connection between two cities through the global sister-city network."),
      selectizeInput("city_a", "Starting City", choices = NULL, options = list(placeholder = 'Search for a city...')),
      selectizeInput("city_b", "Destination City", choices = NULL, options = list(placeholder = 'Search for a city...')),
      hr(),
      uiOutput("path_info")
    ),
    mainPanel(
      leafletOutput("map", height = "600px")
    )
  )
)

server <- function(input, output, session) {
  # Server-side selectize handles the large list of cities efficiently
  updateSelectizeInput(session, "city_a", choices = city_choices, server = TRUE)
  updateSelectizeInput(session, "city_b", choices = city_choices, server = TRUE)
  
  # Reactive to find the shortest path using igraph
  path_ids <- reactive({
    req(input$city_a, input$city_b)
    if (input$city_a == input$city_b) return(input$city_a)
    
    # Verify nodes exist in the graph
    if (!(input$city_a %in% V(links_graph)$name) || !(input$city_b %in% V(links_graph)$name)) {
      return(NULL)
    }
    
    # Find shortest path
    res <- shortest_paths(links_graph, from = input$city_a, to = input$city_b)
    ids <- names(res$vpath[[1]])
    
    if (length(ids) == 0) return(NULL)
    ids
  })
  
  # Display the text-based list of cities in the chain
  output$path_info <- renderUI({
    ids <- path_ids()
    if (is.null(ids)) {
      return(helpText("No connection found between these cities."))
    }
    
    path_names <- cities$name[match(ids, cities$id)]
    tagList(
      h4(paste("Chain Length:", length(ids))),
      tags$ol(
        lapply(path_names, tags$li)
      )
    )
  })
  
  # Render the leaflet map
  output$map <- renderLeaflet({
    ids <- path_ids()
    
    m <- leaflet() %>%
      addProviderTiles(providers$CartoDB.Positron)
    
    if (!is.null(ids)) {
      # Extract coordinates and info in the exact path order
      path_df <- cities[match(ids, cities$id), ]
      
      m <- m %>%
        addMarkers(
          data = path_df, 
          lng = ~lng, lat = ~lat,
          label = ~paste0(name, ", ", country),
          popup = ~paste0("<b>", name, "</b><br>", country)
        ) %>%
        addPolylines(
          lng = path_df$lng, 
          lat = path_df$lat, 
          color = "#e74c3c", 
          weight = 3, 
          opacity = 0.8
        )
      
      # Adjust viewport to show the entire path
      if (nrow(path_df) > 1) {
        m <- m %>% fitBounds(
          lng1 = min(path_df$lng), lat1 = min(path_df$lat),
          lng2 = max(path_df$lng), lat2 = max(path_df$lat)
        )
      } else {
        m <- m %>% setView(lng = path_df$lng[1], lat = path_df$lat[1], zoom = 6)
      }
    }
    m
  })
}

shinyApp(ui, server)
```
