---
title: "TidyTuesday: Sister Cities"
date: "2026-05-12"
format:
  html:
    resources:
      - cities.csv
      - links.csv
categories: ["data science", "rstats", "tidytuesday"]
---

## Introduction

This week's TidyTuesday explores the connections between sister cities across the globe. Unlike previous posts, this one is going to use shiny to create an interactive app.

In particular, I'm interested in identifying chains of sister cities that connect two countries.

### Setup

```{r setup, message=FALSE}
library(tidyverse)
library(shiny)
```

### Loading the Data

```{r load-data, message=FALSE}
# tuesdata <- tidytuesdayR::tt_load('2026-05-12')
# cities   <- tuesdata$cities |>
#   write_csv("cities.csv")
# links    <- tuesdata$links |>
#   write_csv("links.csv")

cities <- read_csv("cities.csv")
links  <- read_csv("links.csv")

```

### Distance chain finding algorithm 

To find the connection between two cities, we can treat the sister city relationships as an undirected graph. The following function uses the `igraph` package to find the the number of links in the shortest path between any two city codes. I'm also creating a tall version of this for different kinds of visualizations.

```{r find-chain}
library(igraph, warn.conflicts = FALSE)

links_graph <- graph_from_data_frame(links, directed = FALSE)
distances <- distances(links_graph)

distances <- as_tibble(distances)

distances_tall <- distances |>
  pivot_longer(everything(), names_to = "to", values_to = "distance") |>
  mutate(from = rep(names(distances), ncol(distances)))

```

```{r most-connected-cities}

number_of_connections <- distances_tall |>
  group_by(from) |>
  summarise(total_connections = sum(distance > 0 & distance != Inf))

ggplot(number_of_connections |>
  mutate(total_connections = as.factor(total_connections)), aes(x = total_connections)) +
  geom_histogram(stat = "count") 
```

It looks like there is one giant web of 4552 sister cities and a cluster of isolated smaller webs.

```{r average-connections}

ggplot(distances_tall, aes(x = distance)) +
  geom_histogram()

```

Let's connect the city ids to the remaining data and see if we can identify some additional patterns

```{r city-data}

number_of_connections_detail <- number_of_connections |>
  left_join(cities, by = c("from" = "id")) |>
  mutate(is_isolated = total_connections != 4552)

connections_formula_location <- is_isolated ~  lng + lat 
number_of_connections_detail |>
  glm(connections_formula_location, family = "binomial", data = _ ) |>
  summary()
```

It looks like the further South you go, the more likely you are to be part of the giant network of cities. Taking this a step further, let's plot all the latitudes and longitudes and categorize them by the number of connections.

```{r city-map}

ggplot(data = number_of_connections_detail |> mutate(total_connections = as.factor(total_connections)), aes(x = lng, y = lat, color = total_connections)) +
  geom_point() +
  theme(legend.position = "bottom") +
  scale_color_brewer(palette = "Set1")

```

Instead of just highlighting the isolated cities, let's actually connect them.

```{r city-connections}

links_detail <- links |>
  left_join(cities |> select(id, lng, lat), by = c("source" = "id")) |>
  rename(from_lng = lng, from_lat = lat) |>
  left_join(cities |> select(id, lng, lat), by = c("target" = "id")) |>
  rename(to_lng = lng, to_lat = lat) |>
  left_join(number_of_connections_detail |> select(from, is_isolated, total_connections),  by= c("source" = "from"))

ggplot(data = links_detail) +
  geom_point(aes(x = from_lng, y = from_lat), color = "grey", alpha = 0.2) +
  geom_segment(data = links_detail |> filter(total_connections <=3), mapping = aes(x = from_lng, y = from_lat, xend = to_lng, yend = to_lat, color = as.factor(total_connections))) +
  theme(legend.position = "bottom")

```

It looks like a mess at the top, but clearly being in the south means you are way less likely to have a small network of sister cities.
