---
title: "Tidy Tuesday: Predicting Castles, Fortresses, and Palaces"
description: "Using tidymodels to classify historical sites based on their age and location."
author: "Mark"
date: "2026-09-01"
categories: [R, TidyTuesday, tidymodels, machine-learning]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This week's [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday) dataset features world castles, fortresses, palaces, and ruins. These historical structures span centuries and continents. In this post, we'll build a predictive model to classify the type of site using its latitude and longitude and the year it was constructed or first documented. Both geography and year help us pinpoint the specific civilization as well as whether we are at nearer the rise or collapse of that civilization and therefore, whether it is likely to be a ruin.

We will use the `tidymodels` framework in R to build a random forest classifier.

## Loading the Data

Let's start by loading the necessary libraries and the dataset.

```{r}
library(tidyverse)
library(tidymodels)

theme_set(theme_light())

# Load the data
# tuesdata <- tidytuesdayR::tt_load('2026-09-01')
# sites <- tuesdata$world_castles
# write.csv(sites, "sites.csv", row.names = FALSE)

sites <- read.csv("sites.csv") |>
  select(qid, name, category, year, lat, lon)

head(sites)

```

## Exploratory Data Analysis

Before building the model, let's visualize the distribution of our historical sites.

```{r}
ggplot(sites, aes(x = lon, y = lat, color = category)) +
  geom_point(alpha = 0.6) +
  labs(title = "Geographic Distribution of Historical Sites",
       x = "Longitude",
       y = "Latitude",
       color = "Site Type") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

It looks like there is some geographical concentration of various site types.

And let's look at the distribution of the construction year by site type.

```{r}
ggplot(sites, aes(x = year)) +
  geom_density(alpha = 1) +
  labs(title = "Age Distribution of Historical Sites",
       x = "Year",
       y = "Density") +
  facet_grid(rows = vars(category)) + 
  theme(axis.text.y = element_blank())

```

It looks like castles and ruins tend to older than fortresses and palaces.

## Building the Model with tidymodels

We'll predict `category` using `year`, `lat`, and `lon`.

### Data Splitting and Resampling

```{r}
set.seed(42)
site_split <- initial_split(sites, strata = category)
site_train <- training(site_split)
site_test <- testing(site_split)
site_folds <- vfold_cv(site_train, strata = category)
```

### Creating the Recipe and Model Specification

Our recipe is simple: we'll predict `category` using all other variables. We'll use a random forest model.

```{r}
site_rec <- recipe(category ~ year + lat + lon, data = site_train)

rf_spec <- rand_forest(trees = 500, min_n = tune(), mtry = tune()) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")

site_wf <- workflow() |>
  add_recipe(site_rec) |>
  add_model(rf_spec)
```

### Tuning the Model

Let's tune the hyperparameters of our random forest.

```{r}
#| eval: true
rf_res <- tune_grid(
  site_wf,
  resamples = site_folds,
  grid = 10,
  control = control_grid(save_pred = TRUE)
)

# Select the best model based on accuracy
best_rf <- select_best(rf_res, metric = "accuracy")

# Finalize the workflow
final_wf <- finalize_workflow(site_wf, best_rf)
```

```{r}
#| echo: false
# # Providing a mock final workflow for rendering purposes
# final_wf <- workflow() |>
#   add_recipe(site_rec) |>
#   add_model(rand_forest(trees = 500, min_n = 5, mtry = 2) |> set_engine("ranger") |> set_mode("classification"))
```

### Final Evaluation

Now, let's fit our final model to the training data and evaluate it on the testing set.

```{r}
final_fit <- last_fit(final_wf, site_split)

collect_metrics(final_fit)
```

Let's visualize the confusion matrix to see where our model succeeds and where it gets confused.

```{r}
collect_predictions(final_fit) |>
  conf_mat(category, .pred_class) |>
  autoplot(type = "heatmap")
```

## Conclusion

In this post, we demonstrated how to use the `tidymodels` framework to classify historical sites based on their location and age. While our model might benefit from additional historical features, it shows how easily geographic and temporal data can be modeled in R.

Happy Tidy Tuesday!
